---
title: "Assignment 7: Build & Evaluate Unsupervised Learning Models"
subtitle: "SoarDDDS8555-6 - Main Controller"
author: "Soar, Donald"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
fontsize: 11pt
geometry: margin=1in
---


```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)

library(reticulate)

python <- "C:/Users/Soar/Documents/.virtualenvs/SoarDDDS8555-6/Scripts/python.exe"

use_python(
  python,
  required = TRUE
)

```



```{r}


dir.create("data", showWarnings = FALSE)

dir.create(
  "scripts",
  showWarnings = FALSE
)

dir.create(
  "output/tables",
  recursive = TRUE,
  showWarnings = FALSE
)

dir.create(
  "output/figures",
  recursive = TRUE,
  showWarnings = FALSE
)

```



```{r}
reticulate::source_python(
  "scripts/assignment_07_01_inspect_wine_data.py"
)

```



```{r}
reticulate::source_python(
   "scripts/assignment_07_02_pca.py"
)

```



```{r}
reticulate::source_python(
  "scripts/assignment_07_03_kmeans.py"
)

```


```{r}

# print(read.csv("output/tables/kmeans_final_cluster_sizes.csv"))

```



```{r}
reticulate::source_python(
   "scripts/assignment_07_04_hierarchical.py"
)

```




```{r}
reticulate::source_python(
  "scripts/assignment_07_05_cluster_profiles.py"
)

```



```{r}
reticulate::source_python(
  "scripts/assignment_07_applied_question_9_usa_rrests.py"
)

```



```{r}

rmarkdown::render(
   "SoarDDDS8555-7-Report.Rmd",
   output_format = "html_document"
)

```



```{r}

rmarkdown::render(
 "SoarDDDS8555-7-Report.Rmd",
  output_format = "pdf_document"
)

```

# References

Barth, J., Katumullage, D., Yang, C., & Cao, J. (2021). Classification of wines using principal component analysis. Journal of Wine Economics, 16(2), 181–198. https://doi.org/10.1017/jwe.2021.1

Jolliffe, I. T., & Cadima, J. (2016). Principal component analysis: A review and recent developments. Philosophical Transactions of the Royal Society A: Mathematical, Physical and Engineering Sciences, 374(2065), 20150202. https://doi.org/10.1098/rsta.2015.0202

MacQueen, J. (1967). Some methods for classification and analysis of multivariate observations. In Proceedings of the Fifth Berkeley Symposium on Mathematical Statistics and Probability (Vol. 1, pp. 281–297). University of California Press. https://digicoll.lib.berkeley.edu/record/113015
